Hexagonal binning algorithm

### 2024-05-13 supplementary/hexagon_binning_algorithm.qmd
Supplementary for hexbin algorithm

```{r}
#| echo: false
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(ggplot2)
library(dplyr)

@@

The 2D hexagon grid is defined by the number of bins in each direction $(b_1, b_2)$, as given by the centroids of each hexagon $(c_1, c_2)$ and the lower left position where the grid starts at $(s_1, s_2)$, which correspond to the lowest left centroid. The values of $s_i$ need to be below their respective minimum variable values, and could be a full bin lower, to allow a buffer corresponding to a distance between centroids in each direction ($a_1$ and $a_2$) around the data. The values of $b_i$ are variables to be computed that define the reduction in size of the data ($n$ to $m$).

The value for $b_2$ is computed by fixing $b_1$. Also, according to hexagonal ratio $a_2 = \frac{\sqrt{3}}{2}a_1$.

The first step of this process is to find the optimal number of bins along the y-axis ($b_1$) at the minimum width of the hexagon ($a_1$) that satisfies the following constraints.

@@

which given $s_1 < 0$ and $a_1 > 0$ will *always* be true. The same logic follows for @eq-equation4 and substituting $a_2 = \frac{\sqrt{3}}{2}a_1$ to @eq-equation4 can be written as,

$$
a_1 > \frac{4}{\sqrt{3}}s_2.
$$
Also, substituting $a_2 = \frac{\sqrt{3}}{2}a_1$ and rearranging @eq-equation5 gives:

$$
a_1 > \frac{2(r_2 - s_2)}{\sqrt{3}(b_2 - 1)}.
$${#eq-equation6}

@@

$$
a_1 > \frac{(1-s_1)}{(b_1 - 1)}.
$${#eq-equation7}

@@

$$
\frac{2(r_2 - s_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1-s_1)}{(b_1 - 1)}.
$$
After rearranging this,

$$
b_2 = 1 + \frac{2(r_2 - s_2)(b_1 - 1)}{\sqrt{3}(1-s_1)}
$$

@@

$$
b_2 = \Big\lceil1 +\frac{2(r_2 - s_2)(b_1 - 1)}{\sqrt{3}(1-s_1)}\Big\rceil.
$$ {#eq-equation8}

@@

$$
\frac{(1-s_1)}{(b_1 - 1)} < \frac{2(r_2 - s_2)}{\sqrt{3}(b_2 - 1)}.
$$
Rearranging this gives,

$$
r_2 > \frac{\sqrt{3}(b_2 - 1)(1-s_1)}{2(b_1 - 1)} + s_2.
$$
Therefore, if this equality is true, then $a_1 = \frac{(1-s_1)}{(b_1 - 1)}$, otherwise, $a_1 = \frac{2(r_2 - s_2)}{\sqrt{3}(b_2 - 1)}$.

```{r, echo=FALSE}
# bin1: Number of bins along the x-axis
# buffer: proportion of buffer should between 0 and 1
# r2: range of embedding 2
calc_bins_y <- function(bin1 = 2, s1 = -0.1, s2 = -0.1, r2) {

  ## Need to have a check where bin2 greater than 2
  if (bin1 < 2) {
    stop("Number of bins along the x-axis at least should be 2.")
  }

  ## To compute the number of bins along the x-axis
  bin2 <- ceiling(1 + (2 * (r2-s2) * (bin1 - 1))/(sqrt(3) * (1-s1)))

  ## Validating and compute horizontal spacing
  check_factor <- (sqrt(3) * (bin2 - 1) * (1-s1))/(2 * (bin1 - 1)) + s2

  if (r2 > check_factor) {

    a1 <- (1-s1)/(bin1 - 1)

  } else {

    a1 <- (2 * (r2-s2))/(sqrt(3) * (bin2 - 1))

  }

  return(list(bin2 = bin2, a1 = a1))
}
```

```{r, echo=FALSE}
gen_centroids <- function(bin1, s1 = -0.1, s2 = -0.1, r2){

  ## To compute hexagonal configurations
  bin_obj <- calc_bins_y(bin1 = bin1, s1 = s1, s2 = s2, r2 = r2)

  # To obtain the bins along the y-axis
  bin2 <- bin_obj$bin2

  # To obtain the width of the hexagon
  a1 <- bin_obj$a1

  # Generate x-coordinate of centroids for odd rows
  c_x_vec_odd <- seq(s1, (bin1 - 1) * a1, by = a1)

  # Generate x-coordinate of centroids for even rows
  c_x_vec_even <- c_x_vec_odd + a1/2
  c_x_vec <- c(c_x_vec_odd, c_x_vec_even)

  # To compute vertical spacing
  a2 <- sqrt(3) * a1/2

  # Generate y-coordinate of centroids
  c_y_vec <- seq(s2, (bin2 - 1) * a2, by = a2)
  c_y <- rep(c_y_vec, each = bin1)

  ## Do the number of belongs y axis is even or odd and adjust the x-coordinates
  if ((bin2 %% 2) == 0) {

    c_x <- rep(c_x_vec, bin2/2)

  } else {

    if ((ceiling(bin2/2) %% 2) == 0) {

      c_x <- append(rep(c_x_vec, floor(bin2/2)), c_x_vec_odd)

    } else{

      c_x <- append(rep(c_x_vec, floor(bin2/2)), c_x_vec_even)

    }

  }

  centroid_df <- tibble::tibble(hexID = 1:length(c_x), c_x = c_x, c_y = c_y)

  return(centroid_df)

}
```

```{r, echo=FALSE}
all_centroids_df <- gen_centroids(bin1 = 4, s1 = -0.1, s2 = -0.1, r2 = 2.019414)
```

```{r, echo=FALSE}
gen_hex_coord <- function(centroids_df, a1){

  ## Obtain centroid info
  hex_ids <- centroids_df$hexID
  c_x_vec <- centroids_df$c_x
  c_y_vec <- centroids_df$c_y

  ## To compute vertical spacing factor
  vs_factor <- a1/(2 * sqrt(3))

  dx <- a1/2
  dy <- a1/sqrt(3)

  ## Assign coordinates for 6 directions
  x_add_factor <- c(0, -dx, -dx, 0, dx, dx)
  y_add_factor <- c(dy, vs_factor, -vs_factor, -dy, -vs_factor, vs_factor)

  ## Initialize vectors to store hexagonal coordinates
  hex_poly_id <- integer(0)
  x <- numeric(0)
  y <- numeric(0)

  for (hb_id in hex_ids) {

    ## Since each hexagon has 6 coordinates
    hexID_rep <- rep(hex_ids[hb_id], each = 6)
    c_x_rep <- rep(c_x_vec[hb_id], each = 6)
    c_y_rep <- rep(c_y_vec[hb_id], each = 6)

    ## Generate the 6 coordinates
    x_spec <- c_x_rep + x_add_factor
    y_spec <- c_y_rep + y_add_factor

    ## Append to existing vectors
    x <- append(x, x_spec)
    y <- append(y, y_spec)
    hex_poly_id <- append(hex_poly_id, hexID_rep)

  }

  hex_coord_df <- tibble::tibble(hex_poly_id = hex_poly_id, x = x, y = y)

  return(hex_coord_df)
}
```

```{r, echo=FALSE}
hex_grid <- gen_hex_coord(all_centroids_df, a1 = 0.3496126)
```

```{r, echo=FALSE}
ggplot(data = hex_grid, aes(x = x, y = y)) + 
    geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
    geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "#31a354") +
    geom_point(data = all_centroids_df |> dplyr::filter(hexID == 1), aes(x = c_x, y = c_y), color = "#08519c") + ## starting point
    geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 2.019414), fill = "white", color = "#feb24c", alpha = 0, linewidth = 1) +
    geom_segment(x = 0, y = -0.25, xend = 1, yend = -0.25,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "#f03b20")+
    annotate("text", x=0.5, y=-0.35, label= "r1", color = "#f03b20") +
    geom_segment(x = -0.25, y = 0, xend = -0.25, yend = 2.019414,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "#f03b20")+ 
    annotate("text", x=-0.35, y=1, label= "r2", color = "#f03b20") +
    geom_segment(x = -0.1, y = -0.15, xend = -0.1 + 0.3496126, yend = -0.15,
                   arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a1 = 0.3496126
    annotate("text", x=0.17, y=-0.35, label= "a1", color = "black") +
    geom_segment(x = -0.15, y = -0.1, xend = -0.15, yend = -0.1 + 0.3027734,
                     arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a2 = 0.3027734
    annotate("text", x=-0.35, y=0.15, label= "a2", color = "black") +
    annotate("text", x=-0.15, y=-0.35, label= "(s1, s2)", color = "#08519c") + ## starting point
    coord_fixed()
```

@@

To find the nearest centroid to a $k^{th}$ embedding point $(y_1, y_2)$ among centroids $(C_1^{(k)}, C_2^{(k)}, \ldots, C_b^{(k)})$,

$$
\text{Minimize: }\sqrt{(y_1 - c_{1k})^2 + (y_2 - c_{2k})^2} + \text{centroid\_ID} \times 10^{-3}. 
$$

In this expression, $(c_{1k}, c_{2k})$ represents the coordinates of centroid $C_1^{(k)}$, and $\text{centroid\_ID}$ represents the ID of the centroid. This formulation ensures that if multiple centroids have the same minimum distance, the one with the minimum ID will be selected.

```{r, echo=FALSE}
emb_df <- tibble::tibble(emb1 = 0.774, emb2 = 0.5055) |>
  add_row(emb1 = 0.424, emb2 = 0.5055) |>
  add_row(emb1 = 0.5115, emb2 = 0.3545) |>
  add_row(emb1 = 0.5115, emb2 = 0.6565) |>
  add_row(emb1 = 0.6865, emb2 = 0.3545) |>
  add_row(emb1 = 0.6865, emb2 = 0.6565) |>
  add_row(emb1 = 0.599, emb2 = 0.304) |>
  add_row(emb1 = 0.599, emb2 = 0.707) |>
  add_row(emb1 = 0.424, emb2 = 0.405) |>
  add_row(emb1 = 0.424, emb2 = 0.606) |>
  add_row(emb1 = 0.774, emb2 = 0.405) |>
  add_row(emb1 = 0.774, emb2 = 0.606)

ggplot(data = hex_grid, aes(x = x, y = y)) + 
    geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
    geom_point(data = emb_df, aes(x = emb1, y = emb2), color = "#253494") +
    geom_segment(x = 0.774, y = 0.5055, xend = 0.7, yend = 0.5055,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.424, y = 0.5055, xend = 0.35, yend = 0.5055,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.599, y = 0.707, xend = 0.599, yend = 0.6,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.774, y = 0.405, xend = 0.774, yend = 0.3,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.774, y = 0.606, xend = 0.7, yend = 0.55,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.6865, y = 0.6565, xend = 0.62, yend = 0.6,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.6865, y = 0.3545, xend = 0.6865, yend = 0.25,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black")

>>>>

```{r}
#| echo: false
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)
```

```{r}
library(ggplot2)
library(dplyr)
library(quollr)
library(readr)

@@

Given range of embedding component, $r_2$, number of bins along the x-axis, $b_1$, and buffer proportion, $q$, hexagonal starting point coordinates, $s_1 = -q$, and $s_2 = -q \times r_2$. The purpose is to find width of the hexagon. $a_1$, and number of bins along the y-axis, $b_2$.

```{r}
# umap_scurve <- read_rds(file = "data/s_curve/s_curve_umap.rds") 
# 
# scurve_scaled_obj <- gen_scaled_data(
#   data = umap_scurve)
# 
# umap_scurve_scaled <- scurve_scaled_obj$scaled_nldr
# lim1 <- scurve_scaled_obj$lim1
# lim2 <- scurve_scaled_obj$lim2
# r2 <- diff(lim2)/diff(lim1)
# 
# a1 <- calc_bins_y(bin1 = 6, q = 0.1, r2 = r2)$a1
r2 <- 1.643542
a1 <- 0.2087578
```

```{r}
all_centroids_df <- gen_centroids(bin1 = 6, q = 0.1, r2 = r2)
```

```{r}
hex_grid <- gen_hex_coord(all_centroids_df, a1 = a1)
```

```{r}
hex_param_vis <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
    geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
    geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "#31a354") +
    geom_point(data = all_centroids_df |> dplyr::filter(hexID == 1), aes(x = c_x, y = c_y), color = "black") + ## starting point
    geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = r2), fill = "white", color = "black", alpha = 0, linewidth = 1) +
    geom_segment(x = 0, y = -0.35, xend = 1, yend = -0.35,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+
    annotate("text", x=0.5, y=-0.45, label= "r1", color = "black") +
    geom_segment(x = -0.25, y = 0, xend = -0.25, yend = r2,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ 
    annotate("text", x=-0.35, y=1, label= "r2", color = "black") +
    geom_segment(x = -0.1 + 0.2087578, y = -0.15, xend = -0.1 + 0.2087578*2, yend = -0.15,
                   arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a1 = 0.2087578
    annotate("text", x=0.2, y=-0.25, label= "a1", color = "black") +
    geom_segment(x = -0.15, y = -0.1*r2 + 0.1807896*2, xend = -0.15, yend = -0.1*r2 + 0.1807896*3,
                     arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a2 = 0.1807896
    annotate("text", x=-0.15, y=0.45, label= "a2", color = "black") +
    annotate("text", x=-0.2, y=-0.35, label= "(s1, s2)", color = "black") + ## starting point
  coord_equal()
```

```{r}
#| label: fig-hex-param
#| fig-cap: "Hexagon binning parameters."

hex_param_vis
```

Geometric arguments give rise to the following constraints.

@@

which given $s_1 < 0$ and $a_1 > 0$ will *always* be true. The same logic follows for @eq-equation4 and substituting $a_2 = \frac{\sqrt{3}}{2}a_1$, and $s_2 = -q \times r_2$ to @eq-equation4 can be written as,

$$
a_1 > -\frac{4}{\sqrt{3}}qr_2
$$ 
Also, substituting $a_2 = \frac{\sqrt{3}}{2}a_1$, $s_2 = -q \times r_2$ and rearranging @eq-equation5 gives:

$$
a_1 > \frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)}.
$$ {#eq-equation6}

@@

$$
a_1 > \frac{(1 + q)}{(b_1 - 1)}.
$$ {#eq-equation7}

@@

$$
\frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1 + q)}{(b_1 - 1)}.
$$ After rearranging this,

$$
b_2 = 1 + \frac{2r_2(b_1 - 1)}{\sqrt{3}}
$$

@@

$$
b_2 = \Big\lceil1 +\frac{2r_2(b_1 - 1)}{\sqrt{3}}\Big\rceil.
$$ {#eq-equation8}

@@

$$
\frac{(1 + q)}{(b_1 - 1)} < \frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)}.
$$

Rearranging this gives,

$$
r_2 > \frac{\sqrt{3}(b_2 - 1)}{2(b_1 - 1)}.
$$

Therefore, if this equality is true, then $a_1 = \frac{(1+q)}{(b_1 - 1)}$, otherwise, $a_1 = \frac{2r_2(1+q)}{\sqrt{3}(b_2 - 1)}$.

@@

```{r}
emb_df <- tibble::tibble(emb1 = 0.774, emb2 = 0.5055) |>
  add_row(emb1 = 0.424, emb2 = 0.5055) |>
  add_row(emb1 = 0.5115, emb2 = 0.3545) |>
  add_row(emb1 = 0.5115, emb2 = 0.6565) |>
  add_row(emb1 = 0.6865, emb2 = 0.3545) |>
  add_row(emb1 = 0.6865, emb2 = 0.6565) |>
  add_row(emb1 = 0.599, emb2 = 0.304) |>
  add_row(emb1 = 0.599, emb2 = 0.707) |>
  add_row(emb1 = 0.424, emb2 = 0.405) |>
  add_row(emb1 = 0.424, emb2 = 0.606) |>
  add_row(emb1 = 0.774, emb2 = 0.405) |>
  add_row(emb1 = 0.774, emb2 = 0.606)

ggplot(data = hex_grid, aes(x = x, y = y)) + 
    geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
    geom_point(data = emb_df, aes(x = emb1, y = emb2), color = "#253494") +
    geom_segment(x = 0.774, y = 0.5055, xend = 0.7, yend = 0.5055,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.424, y = 0.5055, xend = 0.35, yend = 0.5055,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.599, y = 0.707, xend = 0.599, yend = 0.6,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.774, y = 0.405, xend = 0.774, yend = 0.3,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.774, y = 0.606, xend = 0.7, yend = 0.55,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.6865, y = 0.6565, xend = 0.62, yend = 0.6,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
    geom_segment(x = 0.6865, y = 0.3545, xend = 0.6865, yend = 0.25,
                       arrow = arrow(length = unit(0.03, "npc")), color = "black") +
  coord_equal()
